Variational Monte Carlo on Bosonic Systems

### 2018-04-05 project_1/article/article.tex
Add derivation of the exact energy

    \subsection{Non-interacting harmonic oscillators}
        We start by looking at a simple system of non-interacting harmonic
        oscillators, where $a = 0$ and $\beta = 1$. This means that there is no
        hard-shell interaction between the particles and that we use a spherical
        potential trap.  We thus get the trial  wavefunction
        \begin{align}
            \Psi_T(\vf{r})
            = \Phi_T(\vf{r})
            = \prod_{i = 1}^N \exp\para{
                -\alpha |\vf{r}_i|^2
            }.
        \end{align}
        As $a = 0$ the interaction term, $w(\vf{r}_i, \vf{r}_j)$, vanishes and
        the Hamiltonian is given by (in the spherical case)
        \begin{align}
            H &= \sum_{i = 1}^N h(\vf{r}_i)
            = \sum_{i = 1}^N \Biggl(
                -\frac{\hbar^2}{2m}\nabla_i^2
                + \half m \omega^2 r_i^2
            \Biggr).
        \end{align}
        To find the drift force and the local energy we have to compute the
        gradient and the Laplacian of the trial wavefunction. The gradient is
        given by
        \begin{align}
            \nabla_k\Psi_T(\vf{r})
            &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
        \end{align}
        whereas the Laplacian yields
        \begin{align}
            \nabla^2_k\Psi_T(\vf{r})
            &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}).
        \end{align}
        Here $d$ is the dimensionality of the problem determined by $\vf{r}_k
        \in \mathbb{R}^d$. We can thus use the gradient to find an expression
        for the drift force for particle $k$.
        \begin{align}
            \vf{F}_k(\vf{r})
            &= -4\alpha\vf{r}_k.
        \end{align}
        Using the Laplacian we can compute the kinetic term in the expression
        for the local energy. We get
        \begin{align}
            E_L(\vf{r})
            &=
            \sum_{i = 1}^N
            \Biggl(
                -\frac{\hbar^2}{2m}
                \bigl[
                    -2d\alpha + 4\alpha^2 r_i^2
                \bigr]
                + \half m\omega^2 r_i^2
            \Biggr).
        \end{align}
        In natural units, $\hbar = c = 1$, and with the mass set to unity this
        reduces to
        \begin{align}
            E_L(\vf{r})
            &=
            \alpha dN
            + \biggl(
                \half\omega^2
                - 2\alpha^2
            \biggr)
            \sum_{i = 1}^N r_i^2.
            \label{eq:closed_form_natural_units_local_energy}
        \end{align}
        It is worth noting that for $\alpha = \omega/2$\footnote{Note that
        $\alpha$ is required to be positive.} we get a stable value for the
        local energy.
        \begin{align}
            E_L(\vf{r}; \alpha = \omega/2)
            &\equiv
            E_L
            =
            \frac{\omega d N}{2},
            \label{eq:minimum_energy_simple}
        \end{align}
        which turns out to be the exact minimum for the ground state energy of
        this system.

@@

\appendix
\section{Computing the drift force and local energy of the full system}
    Computing the gradient of the wavefunction we get
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Bigl[
            \nabla_k
            \Phi_T(\vf{r})
        \Bigr]
        J(\vf{r})
        + \Phi_T(\vf{r})
        \nabla_k J(\vf{r}).
    \end{align}
    The gradient of the Slater permament for particle $k$ is given by
    \begin{align}
        \nabla_k
        \Phi_T(\vf{r})
        &=
        \nabla_k\phi_k
        \prod_{i \neq k}^N\phi_i
        = \frac{\nabla_k\phi_k}{\phi_k}
        \Phi_T(\vf{r}).
    \end{align}
    The gradient of the Jastrow factor is given by
    \begin{align}
        \nabla_k J(\vf{r})
        &=
        J(\vf{r})
        \nabla_k\sum_{m < n}^N u_{mn} \\
        &= J(\vf{r})
        \Biggl(
            \sum_{m = 1}^{k - 1}\nabla_k u_{mk}
            \sum_{n = k + 1}^N\nabla_k u_{kn}
        \Biggr)
        \\
        &=
        J(\vf{r})
        \sum_{m \neq k}^N\nabla_k u_{km},
    \end{align}
    where the gradient of the interaction term splits the antisymmetric
    sum into two parts. As $r_{ij} = r_{ji}$ we can combine these sums
    into a single sum. This in total yields the gradient
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)
        \Psi_T(\vf{r}).
        \label{eq:gradient_full_wavefunction}
    \end{align}

>>>>

    \subsection{Non-interacting harmonic oscillators}
        We start by looking at a simple system of non-interacting harmonic
        oscillators, where $a = 0$ and $\beta = 1$. This means that there is no
        hard-shell interaction between the particles and that we use a spherical
        potential trap.  We thus get the trial  wavefunction
        \begin{align}
            \Psi_T(\vf{r})
            = \Phi_T(\vf{r})
            = \prod_{i = 1}^N \exp\para{
                -\alpha |\vf{r}_i|^2
            }.
            \label{eq:trial_simple_gaussian}
        \end{align}
        As $a = 0$ the interaction term, $w(\vf{r}_i, \vf{r}_j)$, vanishes and
        the Hamiltonian is given by (in the spherical case)
        \begin{align}
            H &= \sum_{i = 1}^N h(\vf{r}_i)
            = \sum_{i = 1}^N \Biggl(
                -\frac{\hbar^2}{2m}\nabla_i^2
                + \half m \omega^2 r_i^2
            \Biggr).
        \end{align}
        To find the drift force and the local energy we have to compute the
        gradient and the Laplacian of the trial wavefunction. The gradient is
        given by
        \begin{align}
            \nabla_k\Psi_T(\vf{r})
            &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
        \end{align}
        whereas the Laplacian yields
        \begin{align}
            \nabla^2_k\Psi_T(\vf{r})
            &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}).
            \label{eq:laplacian_simple_gaussian}
        \end{align}
        Here $d$ is the dimensionality of the problem determined by $\vf{r}_k
        \in \mathbb{R}^d$. We can thus use the gradient to find an expression
        for the drift force for particle $k$.
        \begin{align}
            \vf{F}_k(\vf{r})
            &= -4\alpha\vf{r}_k.
        \end{align}
        Using the Laplacian we can compute the kinetic term in the expression
        for the local energy. We get
        \begin{align}
            E_L(\vf{r})
            &=
            \sum_{i = 1}^N
            \Biggl(
                -\frac{\hbar^2}{2m}
                \bigl[
                    -2d\alpha + 4\alpha^2 r_i^2
                \bigr]
                + \half m\omega^2 r_i^2
            \Biggr).
        \end{align}
        In natural units, $\hbar = c = 1$, and with the mass set to unity this
        reduces to
        \begin{align}
            E_L(\vf{r})
            &=
            \alpha dN
            + \biggl(
                \half\omega^2
                - 2\alpha^2
            \biggr)
            \sum_{i = 1}^N r_i^2.
            \label{eq:closed_form_natural_units_local_energy}
        \end{align}
        It is worth noting that for $\alpha = \omega/2$\footnote{Note that
        $\alpha$ is required to be positive.} we get a stable value for the
        local energy.
        \begin{align}
            E_L(\vf{r}; \alpha = \omega/2)
            &\equiv
            E_L
            =
            \frac{\omega d N}{2},
            \label{eq:minimum_energy_simple}
        \end{align}
        which turns out to be the exact minimum for the ground state energy of
        this system.

@@

\appendix
\section{Computing the exact energy of the non-interacting system}
    Here we show how to arrive at \autoref{eq:exact_energy}. We need to solve
    two integrals consisting of $N$, $d$-dimensional, Gaussians from
    \autoref{eq:trial_simple_gaussian}. We will label the position in the $j$-th
    coordinate for the $i$-th particle by $x_{ij}$. Starting with the
    normalization factor we get
    \begin{align}
        \bra{\Psi_T}\ket{\Psi_T}
        &=
        \int\dd\vf{r}\abs{\Psi_T(\vf{r})}^2
        \\
        &=
        \prod_{i = 1}^N
        \int\dd\vf{r}_i\exp(-2\alpha\vf{r}_i^2)
        \\
        &=
        \prod_{i = 1}^N
        \prod_{j = 1}^d
        \int\dd x_{ij}\exp(-2\alpha x_{ij}^2)
        \\
        &=
        \prod_{i = 1}^N
        \prod_{j = 1}^d
        \sqrt{\frac{\pi}{2\alpha}}
        = \para{
            \sqrt{\frac{\pi}{2\alpha}}
        }^{dN}.
    \end{align}
    From the Hamiltonian we get two integrals containing $\vf{r}_i^2$ for each
    particle $i$.
    \begin{align}
        \bra{\Psi_T}\vf{r}_i^2\ket{\Psi_T}
        &=
        \int\dd\vf{r}\abs{\Psi_T(\vf{r})}^2\vf{r}_i^2
        \\
        &=
        \int\dd\vf{r}_i\exp(-2\alpha\vf{r}_i^2)\vf{r}_i^2
        \nonumber \\
        &\qquad
        \times
        \prod_{j \neq i}^N
        \int\dd\vf{r}_j\exp(-2\alpha\vf{r}_j^2)
        \\
        &=
        \sum_{l = 1}^d
        \int
        \para{
            \prod_{k = 1}^d
            \dd x_{ik}\exp(-2\alpha x_{ik}^2)
        } x_{il}^2
        \nonumber\\
        &\qquad
        \times
        \para{
            \sqrt{\frac{\pi}{2\alpha}}
        }^{d(N - 1)}
        \\
        &=
        \sum_{l = 1}^d
        \frac{1}{4\alpha}\sqrt{\frac{\pi}{2\alpha}}
        \para{
            \sqrt{\frac{\pi}{2\alpha}}
        }^{d - 1}
        \nonumber\\
        &\qquad
        \times
        \para{
            \sqrt{\frac{\pi}{2\alpha}}
        }^{d(N - 1)}
        \\
        &= \frac{d}{4\alpha}\para{
            \sqrt{\frac{\pi}{2\alpha}}
        }^{dN}.
    \end{align}
    We thus have that
    \begin{align}
        \frac{\bra{\Psi_T}\vf{r}_i^2\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}
        &=
        \frac{d}{4\alpha}.
    \end{align}
    Using \autoref{eq:laplacian_simple_gaussian} we can find the exact energy
    expression we are looking for.
    \begin{align}
        E(\alpha)
        &=
        \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}
        \\
        &=
        \sum_{i = 1}^N
        \brac{
            -\frac{\hbar^2}{2m}
            \brak{
                -2d\alpha
                + 4\alpha^2\frac{d}{4\alpha}
            }
            + \frac{1}{2}m\omega^2\frac{d}{4\alpha}
        }
        \\
        &=
        \para{
            \frac{\alpha\hbar^2}{2m}
            + \frac{m\omega^2}{8\alpha}
        }dN,
    \end{align}
    which is what we wanted to show.

\section{Computing the drift force and local energy of the full system}
    Computing the gradient of the wavefunction we get
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Bigl[
            \nabla_k
            \Phi_T(\vf{r})
        \Bigr]
        J(\vf{r})
        + \Phi_T(\vf{r})
        \nabla_k J(\vf{r}).
    \end{align}
    The gradient of the Slater permament for particle $k$ is given by
    \begin{align}
        \nabla_k
        \Phi_T(\vf{r})
        &=
        \nabla_k\phi_k
        \prod_{i \neq k}^N\phi_i
        = \frac{\nabla_k\phi_k}{\phi_k}
        \Phi_T(\vf{r}).
    \end{align}
    The gradient of the Jastrow factor is given by
    \begin{align}
        \nabla_k J(\vf{r})
        &=
        J(\vf{r})
        \nabla_k\sum_{m < n}^N u_{mn} \\
        &= J(\vf{r})
        \Biggl(
            \sum_{m = 1}^{k - 1}\nabla_k u_{mk}
            \sum_{n = k + 1}^N\nabla_k u_{kn}
        \Biggr)
        \\
        &=
        J(\vf{r})
        \sum_{m \neq k}^N\nabla_k u_{km},
    \end{align}
    where the gradient of the interaction term splits the antisymmetric
    sum into two parts. As $r_{ij} = r_{ji}$ we can combine these sums
    into a single sum. This in total yields the gradient
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &=
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)
        \Psi_T(\vf{r}).
        \label{eq:gradient_full_wavefunction}
    \end{align}
